---
title: "naaccr"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-"
)
```

[![Build status](https://travis-ci.org/WerthPADOH/naaccr.svg?branch=master)](https://travis-ci.org/WerthPADOH/naaccr)

## Summary

The `naaccr` R package enables researchers to easily read and begin analyzing
cancer incidence records stored in the
[North American Association of Central Cancer Registries](https://www.naaccr.org/)
(NAACCR) file format.

## Usage

`naaccr` focuses on two tasks: arranging the records and preparing the fields
for analysis.

### Records

The `naaccr_record` class defines objects which store cancer incidence records.
It inherits from `data.frame`, and for now only makes sure a dataset has a
standard set of columns. While `naaccr_record` has a singular-sounding name, it
can contain multiple records as rows.

The `read_naaccr` function creates a `naaccr_record` object from a
NAACCR-formatted file.

```{r showRecords}
record_file <- system.file(
  "extdata/synthetic-naaccr-18-abstract.txt",
  package = "naaccr"
)
record_lines <- readLines(record_file)
## Marital status and race fields
cat(substr(record_lines[1:5], 206, 216), sep = "\n")
```

```{r readNaaccr}
library(naaccr)

records <- read_naaccr(record_file, version = 18)
records[1:5, c("maritalStatusAtDx", "race1", "race2", "race3")]
```

By default, `read_naaccr` reads all fields defined in a format. For example,
the NAACCR 18 format used above has `r nrow(naaccr_format_18)` fields. Rarely
would an analysis need even 100 fields. By specifying which fields to keep, one
can improve time and memory efficiency.

```{r readKeepColumns}
dim(records)
format(object.size(records))
records_slim <- read_naaccr(
  input       = record_file,
  version     = 18,
  keep_fields = c("ageAtDiagnosis", "countyAtDx", "primarySite")
)
dim(records_slim)
format(object.size(records_slim))
```

Like with most classes, one can create a new `naaccr_record` object with the
function of the same name. The result will have the given columns.

```{r naaccrRecord}
nr <- naaccr_record(
  primarySite = "C010",
  dateOfBirth = "19450521"
)
nr[, c("primarySite", "dateOfBirth")]
```

The `as.naaccr_record` function can transform an existing data frame. It does
require any existing columns to use NAACCR's XML names.

```{r asNaaccrRecord}
prefab <- data.frame(
  ageAtDiagnosis = c(1, 120, 999),
  race1          = c("01", "02", "88")
)
converted <- as.naaccr_record(prefab)
converted[, c("ageAtDiagnosis", "race1")]
```

### Code translation

The NAACCR format uses similar schemes for a lot of fields, and the `naaccr`
package includes functions to help translate them.

`naaccr_boolean` translates "yes/no" fields. By default, it assumes `"0"` stands
for "no", and `"1"` stands for "yes."

```{r naaccrBoolean}
naaccr_boolean(c("0", "1", "2"))
```

Some fields use `"1"` for `FALSE` and `"2"` for `TRUE`. Use the `false_value`
parameter to work with these.

```{r falseValue}
naaccr_boolean(c("0", "1", "2"), false_value = "1")
```

#### Categorical fields

The `naaccr_factor` function translates values using a specific field's category
codes.

```{r naaccrFactor}
naaccr_factor(c("01", "31", "65"), "primaryPayerAtDx")
```

#### Numeric with special missing

Some fields contain primarily continuous or count data but also use special
codes. One name for this type of code is a "sentinel value." The
`naaccr_sentineled` function creates a vector of the `"sentineled"` class.

```{r naaccrSentineled}
rnp <- naaccr_sentineled(c(10, 20, 90, 95, 99, NA), "regionalNodesPositive")
rnp
```

For more on working with `sentineled` vectors, see the documentation for the
`sentinel` package: `help(package="sentinel")`.
