---
title: "naaccr"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-"
)
```

## Summary

The `naaccr` R package enables researchers to easily read and begin analyzing
cancer incidence records stored in the
[North American Association of Central Cancer Registries](https://www.naaccr.org/)
(NAACCR) file format.

## Usage

`naaccr` focuses on two tasks: arranging the records and preparing the fields
for analysis.

### Records

The `naaccr_record` class defines objects which store cancer incidence records.
It inherits from `data.frame`, and for now only makes sure a dataset has a
standard set of columns. While `naaccr_record` has a singular-sounding name, it
can contain multiple records as rows.

The `read_naaccr` function creates a `naaccr_record` object from a
NAACCR-formatted file.

```{r showRecords}
record_file <- system.file(
  "extdata/synthetic-naaccr-18-abstract.txt",
  package = "naaccr"
)
record_lines <- readLines(record_file)
## Marital status and race fields
cat(substr(record_lines[1:5], 206, 216), sep = "\n")
```

```{r readNaaccr}
library(naaccr)

records <- read_naaccr(record_file, version = 18)
records[1:5, c("maritalStatusAtDx", "race1", "race2", "race3")]
```

Like with most classes, one can create a new `naaccr_record` object with the
function of the same name. The result will have all the necessary columns, each
of the correct class. Any columns not provided will be filled with missing
values.

```{r naaccrRecord}
nr <- naaccr_record(
  primarySite         = "C010",
  dateOfBirth         = "19450521"
)
nr[, c("primarySite", "dateOfBirth", "autopsy")]
```

The `as.naaccr_record` function can transform an existing data frame. It does
require any existing columns to use NAACCR's XML names.

```{r asNaaccrRecord}
prefab <- data.frame(
  ageAtDiagnosis = c(1, 120, 999),
  race1          = c("01", "02", "88")
)
as.naaccr_record(prefab)[, c("ageAtDiagnosis", "race1", "anemia")]
```

### Code translation

The NAACCR format uses similar schemes for a lot of fields, and the `naaccr`
package includes functions to help translate them.

`naaccr_boolean` translates "yes/no" fields. By default, it assumes `"0"` stands
for "no", and `"1"` stands for "yes."

```{r naaccrBoolean}
naaccr_boolean(c("0", "1", "2"))
```

Some fields use `"1"` for `FALSE` and `"2"` for `TRUE`. Use the `false_value`
parameter to work with these.

```{r falseValue}
naaccr_boolean(c("0", "1", "2"), false_value = "1")
```

NAACCR's dates follow the `YYYYMMDD` format, which R doesn't recognize. The
`naaccr_date` function parses these strings into `Date` vectors.

```{r naaccrDate}
naaccr_date("20180720")
```

#### Categorical fields

The `naaccr_factor` function translates values using a specific field's category
codes.

```{r naaccrFactor}
naaccr_factor(c("01", "31", "65"), "primaryPayerAtDx")
```

#### Numeric with special missing

Some fields contain primarily continuous or count data but also use special
codes. One name for this type of code is a "sentinel value." The
`naaccr_sentineled` function creates a vector of the `"sentineled"` class.

```{r naaccrSentineled}
rnp <- naaccr_sentineled(c(10, 20, 90, 95, 99, NA), "regionalNodesPositive")
rnp
```

For more on working with `sentineled` vectors, see the documentation for the
`sentinel` package: `help(package="sentinel")`.
